---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cumulcalib

<!-- badges: start -->
[![R-CMD-check](https://github.com/resplab/cumulcalib/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/resplab/cumulcalib/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

cumulcalib provides non-parametric, tuning-parameter-free assessment of model calibration on the cumulative-sum domain, for two settings:

- **Risk prediction models** — `cumulcalib()` assesses the calibration of predicted risks (Sadatsafavi and Petkau 2024, <https://doi.org/10.1002/sim.10138>).
- **Individualized treatment effect (ITE) models** — `cumulcalibITE()` assesses the moderate calibration of predicted treatment benefits using data from a randomized trial (Sadatsafavi et al. 2025, <https://doi.org/10.48550/arXiv.2512.08140>).

The package comes with two tutorials (vignettes), which you can view after installing the package:

``` r
vignette("tutorial", package = "cumulcalib")     # risk prediction models
vignette("tutorialITE", package = "cumulcalib")  # ITE models
```


## Installation

The package can be installed from CRAN:

``` r
install.packages("cumulcalib")
```
You can also install the *development* version, which includes the individualized treatment effect (ITE) functionality, from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes") #this package is necessary to connect to github
remotes::install_github("resplab/cumulcalib")
```

## Example

```{r example}
library(cumulcalib)

set.seed(1)
p <- rbeta(1000, 1,5)
y <- rbinom(1000,1,p)

res <- cumulcalib(y, p)

summary(res)
plot(res)
```

