---
title: "Assessing moderate calibration of individualized treatment effects with cumulcalibITE"
author: "Mohsen Sadatsafavi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moderate calibration of individualized treatment effects (cumulcalibITE)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5, fig.height = 3
)
library(knitr)
```

## Introduction

This is a short companion to the main _cumulcalib_ tutorial, focused on `cumulcalibITE()`, which assesses the **moderate calibration** of predicted _individualized treatment effects_ (ITEs). Where `cumulcalib()` evaluates predicted risks, `cumulcalibITE()` evaluates predicted treatment benefits: the reduction in outcome risk a patient is expected to gain from treatment.

For an ITE model, moderate calibration means that, among individuals with a predicted treatment effect of \(z\), the average true treatment effect is also \(z\). Following the same idea as for risks, this is assessed on the **cumulative-sum domain**: after ordering the validation sample from the smallest to the largest predicted ITE, a standardized process of cumulative prediction errors (the \(\{S\}\) process) is constructed that, under the null hypothesis of moderate calibration, converges to Brownian motion. Inference, visualization, and summary metrics then follow from the properties of Brownian motion. Because treatment benefit is never observed at the individual level (one potential outcome is always counterfactual), the cumulative errors are built from an estimator of the cumulative treatment effect within the ordered sample. See the original paper for the methodology.

### Two approaches: conditional and marginal

Constructing the \(\{S\}\) process requires the (history-)conditional mean and variance of the increments, and these depend on the true baseline (control) risks, which are unknown. The package offers two approaches to resolve this, **both of which assess the same null hypothesis of moderate calibration of the ITEs**:

- **Conditional approach** -- substitute the model's _predicted_ baseline risks for the unknown true risks. This is valid _conditional on_ those predicted risks being calibrated within strata of predicted ITE, and it requires the model to output baseline risks. In `cumulcalibITE()`, this approach is selected by supplying the baseline risk through the `p` argument.

- **Marginal approach** -- replace the cumulative conditional mean and, crucially, the cumulative conditional **variance** of the process with their **marginal** counterparts, estimated from the observed event rates in each treatment arm. This requires no baseline risks and applies to any ITE model (including those that do not predict risks). It is selected by **omitting** `p`. The marginal approach is approximate (heuristic), but in simulations it performs comparably to the conditional approach.

For brevity, this tutorial demonstrates the **marginal approach**; the conditional approach is shown at the end and is obtained simply by adding `p`.

## Setup: an ITE model based on the GUSTO data

We use the GUSTO-I data from the _predtools_ package. The treatment is thrombolytic therapy: accelerated tPA (the active arm, `a = 1`) versus streptokinase (SK, control, `a = 0`); we restrict the data to these two arms. The outcome `y` is 30-day mortality. To obtain a genuine ITE model we include treatment effect modifiers (treatment-by-sex and treatment-by-age interactions), so that the predicted benefit varies across individuals.

```{r}
  library(predtools)
  data(gusto)
  set.seed(1)

  gusto <- gusto[gusto$tx %in% c("tPA", "SK"), ]
  gusto$y      <- gusto$day30
  gusto$a      <- as.numeric(gusto$tx == "tPA")          # 1 = tPA (active), 0 = SK (control)
  gusto$female <- as.numeric(gusto$sex == "female")
  gusto$kill   <- (as.numeric(gusto$Killip) > 1) * 1
  gusto$miloc  <- relevel(factor(gusto$miloc), ref = "Inferior")
  gusto$sbp    <- pmin(gusto$sysbp, 100)
```

As in the main tutorial, we develop the model on the non-US sub-sample and validate it on the US sub-sample. Treatment enters the model as a main effect and in interaction with sex and age (our hypothesized effect modifiers).

```{r}
  dev_data <- gusto[!gusto$regl %in% c(1, 7, 9, 10, 11, 12, 14, 15), ]
  val_data <- gusto[ gusto$regl %in% c(1, 7, 9, 10, 11, 12, 14, 15), ]
  model <- glm(y ~ female + age + miloc + pmi + kill + sbp + pulse + a + a:female + a:age,
               data = dev_data, family = binomial(link = "logit"))
```

The predicted ITE for each validation patient is the difference between their predicted risk under control and under treatment (the predicted reduction in mortality risk):

```{r}
  val_data$pi0 <- predict(model, newdata = transform(val_data, a = 0), type = "response")  # baseline (control) risk
  val_data$pi1 <- predict(model, newdata = transform(val_data, a = 1), type = "response")  # risk under treatment
  val_data$d   <- val_data$pi0 - val_data$pi1                                              # predicted ITE
```

The mean predicted ITE in the validation sample is `r round(mean(val_data$d), 3)`, compared with an observed average treatment effect (ATE) of `r round(mean(val_data$y[val_data$a == 0]) - mean(val_data$y[val_data$a == 1]), 3)`.

## Assessing moderate calibration: the marginal approach

To use the marginal approach we call `cumulcalibITE()` with the outcome `y`, the predicted ITE `d`, and the treatment indicator `a` -- and we **omit** `p`. Following the recommendation in the paper, we use the two-part `"BB"` (bridge) test, which decomposes the assessment into a mean-calibration component and a bridge component and generally has higher power.

```{r}
  library(cumulcalib)
  res <- cumulcalibITE(val_data$y, h = val_data$d, a = val_data$a, method = "BB")
  summary(res)
```

The bridge test reports two components and a unified p-value:

* the **mean-calibration** component (the standardized terminal value `S_n`), a necessary condition for moderate calibration, with p-value `r round(res$by_method$BB$pval_by_component["mean"], 3)`;
* the **bridge** component (the maximum deviation of the standardized walk from the line joining its endpoints), which captures non-linear miscalibration independently of mean calibration, with p-value `r round(res$by_method$BB$pval_by_component["distance"], 3)`;
* the **unified** p-value for moderate calibration (Fisher's combination), `r round(res$by_method$BB$pval, 3)`.

Here the unified p-value is close to, but above, the 0.05 level; the signal comes mainly from the bridge component, indicating mild non-linear miscalibration of the predicted ITEs across their range rather than a systematic mean bias.

## Visual assessment

The returned object is plotted with the same `plot()` method used for risks:

```{r}
  plot(res)
```

The primary X and Y axes show the time and location components of the standardized \(\{S\}\) process; the top X-axis shows the corresponding predicted ITEs, and the right Y-axis shows the scaled (un-standardized) cumulative errors, whose terminal value is `C_n`. The blue markings relate to the mean-calibration component (the terminal value and its significance threshold) and the red markings to the bridge component. One quirk specific to the marginal approach is that, because the marginal variance is not guaranteed to be monotone, the process can occasionally show small negative time increments.

## The conditional approach

If the model produces baseline risks and they can be assumed calibrated within strata of predicted ITE, the conditional approach is available simply by passing those baseline (no-treatment) risks via `p`. Here that is `pi0`:

```{r eval=FALSE}
  # Conditional approach: supply the predicted baseline (control) risk via p
  res_conditional <- cumulcalibITE(val_data$y, h = val_data$d, a = val_data$a,
                                    p = val_data$pi0, method = "BB")
```

The conditional approach has a firmer theoretical footing (a functional central limit theorem can be established for it), but it relies on the calibration of the predicted baseline risks; the marginal approach avoids that assumption at the cost of being heuristic.
