---
title: "Assessing calibration of treatment benefit predictions with cumulcalibITE"
author: "Mohsen Sadatsafavi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration of treatment benefit predictions (cumulcalibITE)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5, fig.height = 3
)
library(knitr)
```

## Introduction

This is a short companion to the main _cumulcalib_ tutorial, focused on `cumulcalibITE()`. While `cumulcalib()` assesses the calibration of predicted _risks_, `cumulcalibITE()` assesses the calibration of predicted _treatment benefits_: the reduction in outcome risk that a patient is expected to gain from treatment (an individualized treatment effect, or ITE, model).

We deliberately reuse the modeling setup from the main tutorial (the GUSTO-I data and the same set of covariates), and simply add the treatment indicator to the model so that it now predicts an individualized treatment benefit. For the underlying methodology, please see the main tutorial (`vignette("tutorial", package = "cumulcalib")`) and the original paper.

## Setup: an ITE model based on the GUSTO data

We again use the GUSTO-I data from the _predtools_ package. The treatment of interest is thrombolytic therapy: we compare accelerated tPA (the active arm, `a = 1`) against streptokinase (SK, the control arm, `a = 0`), and restrict the data to these two arms. As before, the outcome `y` is 30-day mortality and we convert the Killip score to a binary variable.

```{r}
  library(predtools)
  data(gusto)
  set.seed(1)
```

```{r}
  gusto <- gusto[gusto$tx %in% c("tPA", "SK"), ]
  gusto$y    <- gusto$day30
  gusto$a    <- as.numeric(gusto$tx == "tPA")     # 1 = tPA (active), 0 = streptokinase (control)
  gusto$kill <- (as.numeric(gusto$Killip) > 1) * 1
```

Following the main tutorial, we develop the model on the non-US sub-sample and validate it on the US sub-sample. The only change to the model is the addition of the treatment indicator `a` as a covariate, which is what turns this into a treatment-benefit model.

```{r}
  dev_data <- gusto[!gusto$regl %in% c(1, 7, 9, 10, 11, 12, 14, 15), ]
  val_data <- gusto[ gusto$regl %in% c(1, 7, 9, 10, 11, 12, 14, 15), ]
  model <- glm(y ~ a + age + miloc + pmi + kill + pmin(sysbp, 100) + pulse,
               data = dev_data, family = binomial(link = "logit"))
```

To obtain an individualized treatment benefit for each validation patient, we predict their risk under control and under treatment, and take the difference (the predicted reduction in mortality risk):

```{r}
  val_data$p0 <- predict(model, newdata = transform(val_data, a = 0), type = "response")  # risk under control
  val_data$p1 <- predict(model, newdata = transform(val_data, a = 1), type = "response")  # risk under treatment
  val_data$h  <- val_data$p0 - val_data$p1                                                # predicted benefit
```

Here `h` is the predicted benefit, and `p0` (the predicted control risk) serves as the reference risk required by `cumulcalibITE()`. The mean predicted benefit in the validation sample is `r round(mean(val_data$h), 3)`, compared with an observed average treatment effect (ATE) of `r round(mean(val_data$y[val_data$a == 0]) - mean(val_data$y[val_data$a == 1]), 3)`.

## Marginal calibration of the predicted benefit

For a treatment-benefit model, the most fundamental calibration question is whether the _average_ predicted benefit matches the average observed benefit (the observed ATE). This is __marginal__ (also called _mean_) calibration. For brevity, in this short tutorial we suggest performing only this marginal test.

To conduct the marginal test, we call `cumulcalibITE()` with only the outcome `y`, the predicted benefit `h`, and the treatment indicator `a` -- and we do __not__ supply the `p` argument. In this mode the predicted benefit is compared against the observed event rates in the treated and control groups, and no model for the baseline risk is required:

```{r}
  library(cumulcalib)
  res <- cumulcalibITE(val_data$y, h = val_data$h, a = val_data$a)
```

The marginal test is captured by the standardized terminal value of the cumulative process, `S_n`, together with its two-sided p-value:

```{r}
  res$S_n
  res$pval_by_component["mean"]
```

The standardized terminal value is `r round(res$S_n, 2)`, with a marginal p-value of `r round(res$pval_by_component["mean"], 3)`. The average predicted benefit is somewhat smaller than the observed ATE, but this discrepancy does not reach statistical significance at the 0.05 level; the data are compatible with the model being marginally calibrated for the US population.

If one instead wishes to perform the __conditional__ test -- which additionally accounts for the predicted baseline risk -- the baseline risk (the risk _without_ treatment) should be supplied via the `p` argument. In our example this is `p0`, the predicted control risk:

```{r eval=FALSE}
  # Conditional test (not run here): add the baseline (no-treatment) risk via p
  res_conditional <- cumulcalibITE(val_data$y, h = val_data$h, a = val_data$a, p = val_data$p0)
```

## Visual assessment

As with `cumulcalib()`, the returned object can be plotted directly:

```{r}
  plot(res)
```

The blue vertical line marks the terminal value of the random walk -- the marginal (mean) calibration statistic -- and the blue dashed line is its 5% significance threshold. Because the terminal value does not reach the threshold, there is no strong evidence against marginal calibration of the predicted treatment benefit. (The red markings relate to the complementary "distributional" component of calibration, which we do not examine in this short tutorial.)
